% TODO [R2]: Ensure all contribution chapters have consistent closing structure.
% Chapters that HAVE a Conclusion/Discussion section: bfpp, tree2tree, seidr, mimicseq (future work).
% Chapters that are MISSING one: neurogenetic, boptest, heartpole, auto-als, simulators-sota.
% See TODOs placed at the end of each of those chapters.

\chapter{Conclusion: the Future}
\label{ch:conclusion}
\newcommand{\prevchap}{\fpeval{\getrefnumber{ch:conclusion} - 1}}

\section{Highlights of results}

The field of automatic programming has progressed a lot in the timespan of the writing of this thesis\footnote{ensuring a very fun and totally unstressful writing experience}.
Major topics of contention have shifted from "can program synthesis be extended from logic and functional to imperative programming languages?" \cite{polikarpovaStructuringSynthesisHeapmanipulating2019} to "should software developers read the code they deploy?" \cite{andrejkarpathy[@karpathy]TheresNewKind2025}.
Competitive programming is at a historical stage similar to chess in 1996 \cite{pandolfiniKasparovDeepBlue1997}, in which the top 0.2\% of human competitors still outperform the best algorithms \cite{openaiCompetitiveProgrammingLarge2025}.
Part \ref{part:proginduction} traces these developments like a fossil record, from domain-specific languages in chapters \ref{ch:bfpp} and \ref{ch:neurogen} to test-time scaling of language models in chapters \ref{ch:boptest} and \ref{ch:seidr}, advancing the state of the art in both areas.
Methodological proposals advanced in chapters \ref{ch:bfpp}-\ref{ch:seidr} achieve favourable results on standard benchmarks, but perhaps a more important contribution lies in the broader conclusions.
From \emph{expert inspiration} experiments in chapters \ref{ch:bfpp} and \ref{ch:neurogen}, we know that

\begin{highlight}
Import of expert knowledge into a neural and/or evolutionary data-driven system by way of an example program introduced into its training process lets it outperform both the expert and a purely data-driven approach
\end{highlight}

The ability of deep program induction systems to benefit from \emph{importing expert knowledge} (introduced in section \ref{sec:importexport}) was a core motivation for this thesis, and it is gratifying to see this capability progress from a research hypothesis to a research result.

Another major theme of part \ref{part:proginduction} is

\begin{highlight}
Program synthesis challenges are often best solved with hybrid approaches at the intersection of neural, symbolic, and evolutionary methods
\end{highlight}

a verdict that goes against more radical interpretations of \emph{the Bitter Lesson} \cite{BitterLesson2019,traskBitterLessonsBitter2025}.

Part \ref{part:proginduction} concludes with \emph{Synthesize, Execute, Debug, Instruct, Rank} (SEIDR)---a novel program-synthesis framework that introduces testing into language-model-driven program synthesis for the benefit of safety-critical domains such as healthcare.

Part \ref{part:health} advances the state of the art in patient simulators and introduces a suite of benchmarks for reinforcement learning methods in healthcare. It demonstrates that 

\begin{highlight}
in conjunction with a simulator, SEIDR can be used to automatically discover treatment strategies for healthcare settings and often does so \textbf{better than traditional reinforcement learning methods}, while also providing interpretable strategies that enable scientific discovery.
\end{highlight}

At the same time, Auto-ALS (chapter \ref{ch:auto-als})---the most challenging benchmark for Reinforcement Learning from Code Execution in Patient Simulators (RLCEPS) introduced in this thesis---remains unsolved.

\newpage
\section{Future Work}

The negative result on Auto-ALS raises an important question, namely, \emph{what is the missing ingredient?}
What further improvements to RLCEPS are required to validate the method for real-world application in healthcare?
Which questions require further research?
The results in chapters 1--\prevchap{} do not provide a final answer, but do suggest multiple hypotheses:

\paragraph{Neurosymbolic methods}

Chapter \ref{ch:tree2tree} demonstrates how a relatively straightforward combination of language modeling and grammatical parsing can be highly beneficial for downstream tasks.
It also demonstrates how underexplored neurosymbolic program-synthesis methods are: abstract syntax trees are not the only information that can be extracted from source code by means of static analysis, and a tree-based model is not the only way to combine such information with deep learning.

This might be due in part to the popularity of programming copilots (see \ref{sec:sketching}): a copilot built into a developer's interactive environment has to be able to process documents intended as programs that do not necessarily amount to valid programs because of possible mistakes—a challenge for many static-analysis techniques.

In any case, further advances in neurosymbolic methods are important for RLCEPS.

\paragraph{Pre-training curricula}

Many of the methods proposed herein utilize some form of pre-training: prior to facing the patient simulator of choice, a model is trained on a generalist dataset such as a large language model pre-training corpus.
This pipeline could be improved with intermediate steps—something more relevant to the task at hand, yet not as challenging as a patient simulator.
One way to achieve that would be to develop a specialized language model for clinical coding, trained on specialized repositories (and perhaps clinical textbooks), and use it with SEIDR (chapter \ref{ch:seidr}) to tackle a patient simulator.

\paragraph{Protocol digitalisation}

RLCEPS was designed to enable \emph{expert inspiration} in training data-driven models of healthcare (proposed in section \ref{sec:importexport}, operationalized in section \ref{sec:bfpp-ps-model}).
As in chapters \ref{ch:bfpp} and \ref{ch:neurogen}, where an expert-written program is introduced early into the training process, a program written by an expert in intensive care could be introduced as a starting point for solving \emph{MIMIC-SEQ} (chapter \ref{ch:mimicseq}).
This program would describe an accepted clinical protocol for an intensive care ward.

During research on this topic, one limitation became (more) apparent: clinical protocols are not typically represented as code.
To be used in RLCEPS context, a clinical protocol has to be translated from some source material (e.g. a clinical textbook) into code.
This translation is a challenging program synthesis task on its own, particularly due to the complexity of source material: note that some important information in textbooks is presented in diagram format, making it some of the most challenging input formats for deep learning models \cite{hauriletMoQAMultimodalQuestion2019,heTextbookQuestionAnswering2021,jamiesonDeepLearningText2020,jamiesonReviewDeepLearning2024,novakDiagramsTextComputer1993,shiinokiOvercomingVisionLanguage2025}.
Shall it be done manually, it requires the attention of a programming expert as well as an expert in the clinical field to avoid misinterpretation.

Representing clinical protocols in a format amenable to digital analysis and execution in simulators is an important stepping stone to RLCEPS.

\paragraph{Scaling}

Finally, one cannot discount the possibility that the real missing ingredient is computational \textbf{scale}, i.e., applying the same technique for longer with more hardware will solve Auto-ALS and more.
After all, this has been the case for many deep-learning methods, including neural networks themselves \cite{schmidhuberAnnotatedHistoryModern2022}.
With that in mind, all experiments in this thesis were designed to be scaling-friendly.
Chapter \ref{ch:mimicseq}, in particular, lays out a plan for what the next steps in RLCEPS should be, given enough computational resources.

\clearpage
\thispagestyle{empty}
\vspace*{\fill}
\begin{center}
\emph{Finis tantum principium est}
\end{center}
\vspace*{\fill}

